Concurrent Programming in Harmony

### 2021-01-19 book/paper.tex
Latest

\begin{figure}
\begin{code}
\begin{verbatim}
#states 32
Safety Violation
T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
T1: incrementer(0) [6 - 9] { count: 0, done: [ False, False ] }
T2: incrementer(1) [6 -18] { count: 1, done: [ False, True  ] }
T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
Harmony assertion failed: 1
open file:///Users/rvr/github/harmony/harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one thread, named \texttt{\_\_init\_\_},
which initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, it executes instructions 0 through 4 first.
The last instruction is a \texttt{JUMP} instruction that sets the
program counter to 31 (skipping over the code for the methods).
At program counter 5 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a \texttt{Return}
instruction.

@@

\begin{enumerate} \setcounter{enumi}{5}
\item The \texttt{Load} instruction pushes the value of the
\textit{count} variable onto the stack.
\item The \texttt{Push} instruction pushes the constant 1
onto the stack of the thread.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (the value of \textit{count} and 1),
adds them, and pushes the result back onto the stack.
\item The \texttt{Store} instruction pops
a Harmony value (the sum of the \textit{count} variable and 1) and
stores it in the \textit{count} variable.
\end{enumerate}

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/Up1.png}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are three sections.
The top shows the steps to the problematic state.
The bottom left shows the source code and bytecode
of the program.
The bottom right shows the state of each thread after the selected step.
Each row in the stack trace of a thread shows the current program counter,
the method that is being evaluated, and the line of code that is
being evaluated.}
\label{fig:inchtml1}
\end{figure}

Harmony can report a variety of failure types:
\begin{itemize}
\item \texttt{Safety violation}: this means something went wrong with
at least one of the executions of the program that it tried.  This
can include a failing assertion, divide by zero, using an uninitialized
or non-existent variable, dividing a set by an integer, and so on.
It also includes certain infinite loops.
Harmony will print at trace of the shortest bad execution that it found.
\item \texttt{Non-terminating States}: Harmony found one or more states
from which there does not exist an execution such that all threads
terminate.  Harmony will not only print the non-terminating state with
the shortest trace, but also the list of threads
at that state, along with their program counters.
\item \texttt{Stopped States}: Similar to non-terminating states,
these are states in which some threads are left that cannot
terminate.  \autoref{ch:synch} will explain what it means for
a thread to be stopped.
\end{itemize}

In our case, Harmony reports a safety violation, in particular it
reports that the assertion and that
\textit{count} has value 1 instead of 2.
Next it reports an execution that failed this assertion.
The program got to the faulty execution in 5 ``steps.''
The output has five columns:
\begin{enumerate}
\item A thread identifier;
\item The main method and argument of the thread;
\item The sequence of program counters of the HVM instructions that the thread executed;
\item The current program counter of the thread;
\item The contents of the shared memory.
\end{enumerate}

The first step is initialization.
It sets shared variable \textit{count} to 0 and
shared variable \textit{done} to \texttt{[False, False]}.
If we look at the next three steps, we see that:
\begin{itemize}
\item Thread \texttt{incrementer(0)}
executed instructions 5 through 8, loading the value of
\textit{count} but stopping just before storing 1 into \textit{count};
\item Thread \texttt{incrementer(1)}
executed instructions 5 through 15, storing 1 into
into \textit{count} and storing \texttt{True} into \textit{done}[1];
\item Thread \texttt{incrementer(0)}
continues execution, executing instructions 9 through 15
storing value 1 into \textit{count}
and storing \texttt{True} into \textit{done}[0].
\end{itemize}
Finally, thread \texttt{main()} runs and finds the problem.
This makes precise the concurrency issue that we encountered.

Harmony also generates an HTML file that allows exploring more details
of the execution interactively.
Open the suggested HTML file and you should see something like
\autoref{fig:inchtml1}.
If you hover the mouse over a machine instruction, it provides a
brief explanation.

>>>>

\begin{figure}
\begin{code}
\begin{verbatim}
#states 32
Safety Violation
T0: __init__() [0-5,32-40] { count: 0, done: [ False, False ] }
T1: incrementer(0) [ 6- 9] { count: 0, done: [ False, False ] }
T2: incrementer(1) [ 6-18] { count: 1, done: [ False, True  ] }
T1: incrementer(0) [10-28] { count: 1, done: [ True,  True  ] }
Harmony assertion failed: 1
open file:///Users/rvr/github/harmony/harmony.html for more information
\end{verbatim}
\end{code}
\caption{The text output of running Harmony on \autoref{fig:inc}.}
\label{fig:incoutput}
\end{figure}

@@

All instructions are atomically executed.
Most instructions pop values from the stack or push values onto the stack.
At first there is one thread, named \texttt{\_\_init\_\_},
which initializes the state.
It starts executing at instruction 0 and keeps executing until the last
execution in the program.
In this case, it executes instructions 0 through 5 first.
The last instruction is a \texttt{JUMP} instruction that sets the
program counter to 32 (skipping over the code for \texttt{incrementer} method).
At program counter 6 is the code for the \texttt{incrementer} method.
All methods start with a \texttt{Frame} instruction and end with a
\texttt{Return} instruction.

@@

\begin{enumerate} \setcounter{enumi}{6}
\item The \texttt{Load} instruction pushes the value of the
\textit{count} variable onto the stack.
\item The \texttt{Push} instruction pushes the constant 1
onto the stack of the thread.
\item \texttt{2-ary} is a \texttt{+} operation with 2 arguments.
It pops two values from the stack (the value of \textit{count} and 1),
adds them, and pushes the result back onto the stack.
\item The \texttt{Store} instruction pops
a Harmony value (the sum of the \textit{count} variable and 1) and
stores it in the \textit{count} variable.
\end{enumerate}

@@

\begin{figure}
\begin{center}
\includegraphics[width=0.9\textwidth]{figures/Up1.png}
\end{center}
\caption{The
\harmonyref{Up.html}{HTML output of running Harmony on \autoref{fig:inc}}.
There are four sections.
The top shows the steps to the problematic state.
Each step shows a thread identifier, the method the thread is executing,
a ``timeline'' containing a block for each instruction executed, the
program counter of the thread, and the values of the shared variables
at those steps.
The middle section shows the line of code that is being executed.
The bottom left shows the bytecode of the program.
The bottom right shows the state of each thread.
Each row in the stack trace of a thread shows
the method that is being evaluated and the values of its method variables.}
\label{fig:inchtml1}
\end{figure}

Harmony can report a variety of failure types:
\begin{itemize}
\item \texttt{Safety violation}: this means something went wrong with
at least one of the executions of the program that it tried.  This
can include a failing assertion, divide by zero, using an uninitialized
or non-existent variable, dividing a set by an integer, and so on.
It also includes certain infinite loops.
Harmony will print a trace of the shortest bad execution that it found.
\item \texttt{Non-terminating States}: Harmony found one or more states
from which there does not exist an execution such that all threads
terminate.  Harmony will not only print the non-terminating state with
the shortest trace, but also the list of threads
at that state, along with their program counters.
\item \texttt{Stopped States}: Similar to non-terminating states,
these are states in which some threads are left that cannot
terminate.  \autoref{ch:synch} will explain what it means for
a thread to be stopped.
\end{itemize}

In our case, Harmony reports a safety violation, in particular it
reports that the assertion and that
\textit{count} has value 1 instead of 2.
Next it reports an execution that failed this assertion.
The program got to the faulty execution in 5 ``steps.''
The output has five columns:
\begin{enumerate}
\item A thread identifier;
\item The main method and argument of the thread;
\item The sequence of program counters of the HVM instructions that the thread executed;
\item The contents of the shared memory.
\end{enumerate}

The first step is initialization.
It sets shared variable \textit{count} to 0 and
shared variable \textit{done} to \texttt{[False, False]}.
If we look at the next three steps, we see that:
\begin{itemize}
\item Thread \texttt{incrementer(0)}
executed instructions 6 through 9, loading the value of
\textit{count} but stopping just before storing 1 into \textit{count};
\item Thread \texttt{incrementer(1)}
executed instructions 6 through 18, storing 1 into
into \textit{count} and storing \texttt{True} into \textit{done}[1];
\item Thread \texttt{incrementer(0)}
continues execution, executing instructions 10 through 28
storing value 1 into \textit{count} (instruction 10),
storing \texttt{True} into \textit{done}[0] (instructions 11 through 15)
finding that \textit{done}[1] is \texttt{True} (instructions 16 through 21),
and finally detecting that the assertion is violated
(instructions 22 through 28).
\end{itemize}

Harmony also generates an HTML file that allows exploring more details
of the execution interactively.
Open the suggested HTML file and you should see something like
\autoref{fig:inchtml1}.
You can go to any point in the execution by clicking on one of the
blocks in the ``execution timelines.''  Using the left and right arrow
keys you can go forwards and backwards one instruction at a time.
With the down and up arrow keys you can go through the execution one
step at a time.
If you hit enter, it will to go to the next line of Harmony code.
If you hover the mouse over a machine instruction, it provides a
brief explanation.
